---
title: "EDA"
author: "Trever Yoder"
format: html
editor: visual
---

### Data

```{r}
# Load packages
library(ggplot2)
library(dplyr)
```


```{r}
# Read in the data
diabetes_data <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")

# Pick response and variables I will investigate
diabetes_data <- diabetes_data[, c("Diabetes_binary", "BMI", "PhysActivity", "HighBP")] %>%
    mutate(
    Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")),
    PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
    HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes"))
  )
head(selected_data)
```

Our response variable will be *Diabetes_binary* and the 3 selected predictors are: *BMI*, *PhysActivity*, and *HighBP.* Studies have shown strong links between these variables and diabetes outcomes. For example, [this article by Mokdad et al. (2005)](https://diabetesjournals.org/care/article/28/7/1599/27992/Obesity-Inactivity-and-the-Prevalence-of-Diabetes) published in *Diabetes Care* highlights how **obesity and inactivity** contribute significantly to the prevalence of diabetes.This article also mentions blood pressure and BMI.


### Plots

```{r}
# Histogram of BMI
ggplot(diabetes_data, aes(x = BMI)) +
  geom_histogram(binwidth = 2, fill = "steelblue") +
  labs(title = "Distribution of BMI")
```

```{r}
# Bar plot of PhysActivity
ggplot(diabetes_data, aes(x = PhysActivity)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Physical Activity Distribution", x = "Physical Activity in the Last 30 Days")
```



```{r}
# Bar plot of HighBP
ggplot(diabetes_data, aes(x = HighBP)) +
  geom_bar(fill = "firebrick") +
  labs(title = "High Blood Pressure Distribution", x = "High Blood Pressure", y = "Count")
```


```{r}
# BMI vs. Diabetes
ggplot(diabetes_data, aes(x = Diabetes_binary, y = BMI)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "BMI by Diabetes Status", x = "Diabetes Status", y = "BMI")
```

```{r}
# PhysActivity vs. Diabetes
ggplot(diabetes_data, aes(x = PhysActivity, fill = Diabetes_binary)) +
  geom_bar(position = "fill") +
  labs(title = "Diabetes Rate by Physical Activity", x = "Physical Activity", y = "Proportion", fill = "Diabetes")
```



```{r}
# HighBP vs. Diabetes
ggplot(diabetes_data, aes(x = HighBP, fill = Diabetes_binary)) +
  geom_bar(position = "fill") +
  labs(title = "Diabetes Rate by High Blood Pressure", x = "High Blood Pressure", y = "Proportion", fill = "Diabetes")
```







[Click here for the Modeling Page](Modeling.html)
